A full approximation scheme multilevel method \\ for nonlinear variational inequalities

### 2023-12-26 paper/fascd.tex
more edits, through middlle section 4

\noindent The sum ``$\sum_{j>i} \Pi_j u$'' should be read as ``those parts of $u$ which need improvement''.  Tai \cite{Tai2003} also proposes an additive/parallel version of the algorithm, but it is not used here.

The VI problems in \pr{cd-mult} reduce to constrained line searches when the $\cV_i$ are one-dimensional.  If $e_i = w_i - \Pi_i u \in \cV_i$ denotes the correction from $\cK_i$, then the new iterate $w = u + \sum_i e_i$ is computed by adding a correction from each sub\emph{space} $\cV_i$, but so that $w \in \cK$; the corrections preserve admissibility.  Furthermore, if VI problem \eqref{eq:vi} corresponds to convex optimization \eqref{eq:minimization} then \pr{cd-mult} generates monotonically-decreasing objective values.

@@

\pr{cd-mult} is meaningful whether or not we are in the optimization case $f=F'$, and even if $f$ is nonlinear, non-local, or defined only on $\cK$, but its practical and efficient FE implementation for such general operators $f$ seems not to have been addressed in the literature.  In particular, references \cite{GraeserKornhuber2009,Tai2003} only apply the multilevel CD method to the classical obstacle problem, wherein $f=F'$ is linear, local, and defined on all of $\mathcal{V}$.  In Section \ref{sec:vcycle} we will extend multilevel CD iterations to general nonlinear operators by applying the full approximation scheme (FAS) idea of Brandt \cite{Brandt1977}, and then illustrate its application to general operators in Examples \ref{ex:results:plap}--\ref{ex:results:sia}.

From now on we restrict to the case of \emph{box constraints} \cite{BensonMunson2006,FerrisPang1997}, equivalently \emph{bilateral obstacle problems}, over $\mathcal{V}=W^{1,p}(\Omega)$.  The \emph{lower obstacle} $\underline{\gamma}$ and \emph{upper obstacle} $\overline{\gamma}$ are given measurable functions, defined on the closure $\overline{\Omega}$, with values in the extended real line $\tilde\RR = [-\infty,+\infty]$, satisfying the following properties: $\underline{\gamma} < +\infty$, $\overline{\gamma}>-\infty$, and $\underline{\gamma} \le \overline{\gamma}$.  Also suppose $\partial\Omega$ is split into disjoint sets $\partial\Omega = \partial_D \Omega \cup \partial_N \Omega$, with $\partial_D \Omega$ of positive Hausdorff measure, and $\partial_N \Omega$ possibly empty.  Dirichlet data is given by $g_D:\partial_D \Omega \to \RR$, and we assume $\underline{\gamma} \le g_D \le \overline{\gamma}$ on $\partial_D \Omega$.  Now define
\begin{equation}
\cK = \left\{w\,:\,\underline{\gamma} \le w \le \overline{\gamma} \, \text{ and }\, w\big|_{\partial_D \Omega} = g_D\right\} \subset \cV =W^{1,p}(\Omega), \label{eq:originalconstraintset}
\end{equation}
the closed and convex \emph{constraint set}; this treats inequality constraints and Dirichlet conditions in a unified manner.  Dirichlet values are in the trace sense \cite{Evans2010}, and $g_D$ is assumed to be as regular as needed.  Any Neumann conditions over $\partial_N \Omega$ are imposed weakly by the operator $f$, via an integral over $\partial_N\Omega$.  We will numerically solve VI problems \eqref{eq:vi} over constraint sets \eqref{eq:originalconstraintset}.

@@

We seek to rapidly compute the approximate solution of box-constrained VI problems \eqref{eq:vi} and \eqref{eq:originalconstraintset} using finite elements (FE).  In this Section a nested multilevel $P_1$ or $Q_1$ element \cite{Elmanetal2014} approximation to \eqref{eq:vi} will be constructed by refinement of a coarse mesh.  The mesh-refinement hierarchy permits the construction of a multilevel CD of the finest-level constraint set.

The mesh hierarchy and FE spaces can be constructed by standard nested refinement.  Let $\Omega \subset \RR^d$ be an open bounded polygon and suppose $\mathcal{T}^0$ is the \emph{coarsest mesh}, a finite set of non-overlapping, nondegenerate cells with union equal to $\overline{\Omega}$ and maximum cell diameter (mesh size) $h_0>0$.  We assume $\mathcal{T}^0$ is conforming, with no hanging nodes.  For $j=1,\dots,J$, let $\mathcal{T}^j$ be the uniform refinement of $\mathcal{T}^{j-1}$; each $\mathcal{T}^{j-1}$ cell becomes $2^d$ cells in $\mathcal{T}^j$ with mesh size $h_j = 2^{-j} h_0$.  We call $\mathcal{T}^J$ the \emph{finest-level mesh} and $J+1$ the \emph{number of levels}.  Let $m_j$ be the number of vertices in $\mathcal{T}^j$, the \emph{degrees of freedom} at the $j$th level.

Let $\mathcal{V}^j$ be nested FE spaces over each mesh $\mathcal{T}^j$, with arbitrary boundary values:
\begin{equation}
\mathcal{V}^0 \subset \mathcal{V}^1 \subset \dots \subset \mathcal{V}^J \subset \mathcal{V}=W^{1,p}(\Omega).  \label{eq:fe:nestedspaces}
\end{equation}
As is common in the discretization of variational inequalities, we employ lowest-order continuous Lagrange elements $P_1$ or $Q_1$ on simplices and tensor-product cells respectively\footnote{Keith \& Surowiec have recently proposed a scheme for solving VIs using high-order finite elements~\cite{Keith2023}.}. This choice is convenient because it enjoys \emph{nodal monotonicity}:
\begin{equation}
\bw \ge \bz \quad \implies \quad w \ge z \label{eq:nodalmonotonicity}
\end{equation}
for all $w,z \in \mathcal{V}^j$, where $\bw,\bz \in \RR^{m_j}$ denote vectors of nodal values. In other words, to enforce an inequality constraint, with $P_1$ or $Q_1$ we only enforce it pointwise on the degrees of freedom. For $k\ge 2$ there are functions in the standard bases of $P_k$ and $Q_k$ which take on negative values \cite[Figure 1.7]{Elmanetal2014}, violating \eqref{eq:nodalmonotonicity}.

To construct box constraints on all meshes in the hierarchy, let $\tilde{\mathcal{V}}^j$ denote the set of functions on the vertices of $\mathcal{T}^j$ with values in $\tilde{\RR} = [-\infty,+\infty]$.  The finest-level \emph{lower} and \emph{upper obstacles} $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$ are constructed by suitable interpolation of $\underline{\gamma}$, $\overline{\gamma}$, e.g.~by applying the Cl\'ement interpolation operator~\cite{Carstensen2006}.  Again the obstacles must satisfy $\underline{\gamma}^J < +\infty$, $-\infty < \overline{\gamma}^J$, and $\underline{\gamma}^J \le \overline{\gamma}^J$.  On the finest-level mesh the discrete boundary data $g_D^J \in \mathcal{V}^J$ is assumed to satisfy $g_D^J = g_D$ on $\partial_D \Omega$---this restricts $g_D$ in \eqref{eq:originalconstraintset} to be piecewise-linear---and $\underline{\gamma}^J \le g_D^J \le \overline{\gamma}^J$ on those vertices of $\mathcal{T}^J$ which lie in $\partial_D \Omega$.  Then we define
\begin{equation}
\mathcal{K}^J = \big\{w\,:\,\underline{\gamma}^J \le w \le \overline{\gamma}^J \text{ on vertices of } \mathcal{T}^J, \, \text{ and } \, w|_{\partial_D\Omega} = g_D^J|_{\partial_D\Omega}\big\} \subset \mathcal{V}^J. \label{eq:fe:fineconstraintset}
\end{equation}
This nonempty, closed, and convex subset is the finest-level \emph{constraint set}.

Let $f^J:\mathcal{K}^J \to (\mathcal{V}^J)'$ be a discretization of $f$ in \eqref{eq:vi}.  Note that the functional $\ell^J=\ell$ is defined and continuous over $\mathcal{V}^J$.  We seek the FE solution $u^J \in \mathcal{K}^J$ of the finest-level, finite-dimensional, nonlinear VI
\begin{equation}
\ip{f^J(u^J)}{v-u^J} \ge \ip{\ell^J}{v-u^J} \qquad \text{for all } v\in \cK^J. \label{eq:fe:vi}
\end{equation}
If $f^J$ is continuous and $p$-coercive on $\mathcal{K}^J$, in the sense that \eqref{eq:pcoercive} holds for all $u,v \in \mathcal{K}^J$, then \eqref{eq:fe:vi} is well-posed.

>>>>

\noindent The sum ``$\sum_{j>i} \Pi_j u$'' should be read as ``portions of $u$ needing improvement''.  Tai \cite{Tai2003} also proposes an additive/parallel version of the algorithm, but it is not used here.

The VI problems inside \pr{cd-mult} reduce to constrained line searches when the $\cV_i$ are one-dimensional.  If $e_i = w_i - \Pi_i u \in \cV_i$ denotes the correction from $\cK_i$, then the new iterate $w = u + \sum_i e_i$ is computed by adding a correction from each subspace $\cV_i$, but so that $w \in \cK$; the corrections preserve admissibility.

If VI problem \eqref{eq:vi} corresponds to convex optimization \eqref{eq:minimization} then \pr{cd-mult} generates monotonically-decreasing objective values.

@@

However, \pr{cd-mult} is meaningful whether or not we are in the optimization case $f=F'$, and even if $f$ is nonlinear, non-local, or defined only on $\cK$.  Practical and efficient FE implementation for such general operators $f$ seems not to have been addressed in the literature.  In particular, references \cite{GraeserKornhuber2009,Tai2003} only apply the multilevel CD method to the classical obstacle problem, wherein $f=F'$ is linear, local, and defined on all of $\mathcal{V}$.  In Section \ref{sec:vcycle} we will extend multilevel CD iterations to general nonlinear operators by applying the full approximation scheme (FAS) idea of Brandt \cite{Brandt1977}, and then illustrate its application to general operators in Section \ref{sec:results}.

From now on we restrict to the case of \emph{box constraints} \cite{BensonMunson2006,FerrisPang1997}, also called bilateral obstacle problems, over $\mathcal{V}=W^{1,p}(\Omega)$.  The \emph{lower obstacle} $\underline{\gamma}$ and \emph{upper obstacle} $\overline{\gamma}$ are given measurable functions, defined on the closure $\overline{\Omega}$, with values in the extended real line $\tilde\RR = [-\infty,+\infty]$, satisfying the following properties: $\underline{\gamma} < +\infty$, $\overline{\gamma}>-\infty$, and $\underline{\gamma} \le \overline{\gamma}$.  We also suppose that $\partial\Omega$ is split into disjoint sets $\partial\Omega = \partial_D \Omega \cup \partial_N \Omega$, with $\partial_D \Omega$ of positive Hausdorff measure, and $\partial_N \Omega$ possibly empty.  Dirichlet data is given by $g_D:\partial_D \Omega \to \RR$, and we assume $\underline{\gamma} \le g_D \le \overline{\gamma}$ on $\partial_D \Omega$.  Now define
\begin{equation}
\cK = \left\{w\,:\,\underline{\gamma} \le w \le \overline{\gamma} \, \text{ and }\, w\big|_{\partial_D \Omega} = g_D\right\} \subset \cV =W^{1,p}(\Omega), \label{eq:originalconstraintset}
\end{equation}
the closed and convex \emph{constraint set}.  (Observe that this treats inequality constraints and Dirichlet conditions in a unified manner.)  Dirichlet values are in the trace sense \cite{Evans2010}, and $g_D$ is assumed to be as regular as needed.  Any Neumann conditions over $\partial_N \Omega$ are imposed weakly by the operator $f$, via an integral over $\partial_N\Omega$.

@@

We will approximate solutions of box-constrained VI problems \eqref{eq:vi}, \eqref{eq:originalconstraintset} using finite elements (FE).  In this Section a nested, multilevel, and $P_1$ or $Q_1$ element \cite{Elmanetal2014} mesh hierarchy is constructed by refinement of a coarse mesh.  This hierarchy permits the construction of a multilevel CD of the finest-level constraint set in Section \ref{sec:cdmultilevel}.

The construction is by standard nested refinement as follows.  Let $\Omega \subset \RR^d$ be an open bounded polygon and suppose $\mathcal{T}^0$ is the \emph{coarsest mesh}, a finite set of non-overlapping, nondegenerate cells with union equal to $\overline{\Omega}$ and maximum cell diameter (mesh size) $h_0>0$.  We assume $\mathcal{T}^0$ is conforming, with no hanging nodes.  For $j=1,\dots,J$, let $\mathcal{T}^j$ be the uniform refinement of $\mathcal{T}^{j-1}$; each $\mathcal{T}^{j-1}$ cell becomes $2^d$ cells in $\mathcal{T}^j$ with mesh size $h_j = 2^{-j} h_0$.  We call $\mathcal{T}^J$ the \emph{finest-level mesh} and $J+1$ the \emph{number of levels}.  Let $m_j$ be the number of vertices in $\mathcal{T}^j$, the number of degrees of freedom at the $j$th level.

As is common in the discretization of variational inequalities, we employ lowest-order continuous Lagrange elements $P_1$ or $Q_1$ on simplices and tensor-product cells respectively.\footnote{Keith \& Surowiec have recently proposed a scheme for solving VIs using high-order finite elements~\cite{Keith2023}.}  Let $\mathcal{V}^j$ be such nested FE spaces over each mesh $\mathcal{T}^j$, with arbitrary boundary values:
\begin{equation}
\mathcal{V}^0 \subset \mathcal{V}^1 \subset \dots \subset \mathcal{V}^J \subset \mathcal{V}=W^{1,p}(\Omega).  \label{eq:fe:nestedspaces}
\end{equation}
The choice of $P_1$ or $Q_1$ is convenient because it enjoys \emph{nodal monotonicity}:
\begin{equation}
\bw \ge \bz \quad \implies \quad w \ge z \label{eq:nodalmonotonicity}
\end{equation}
for all $w,z \in \mathcal{V}^j$, where $\bw,\bz \in \RR^{m_j}$ denote vectors of nodal values. In other words, to enforce an inequality constraint over $\cV^j$ we need only enforce it pointwise on the degrees of freedom.  By contrast, for $k\ge 2$ there are functions in the standard bases of $P_k$ and $Q_k$ which take on negative values \cite[Figure 1.7]{Elmanetal2014}, violating \eqref{eq:nodalmonotonicity}.

To construct box constraints on all meshes in the hierarchy, let $\tilde{\mathcal{V}}^j$ denote the set of $\tilde{\RR} = [-\infty,+\infty]$ valued functions on the vertices of $\mathcal{T}^j$.  The finest-level \emph{lower} and \emph{upper obstacles} $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$ are constructed by suitable interpolation of $\underline{\gamma}$, $\overline{\gamma}$, e.g.~by applying the Cl\'ement interpolation operator~\cite{Carstensen2006}.  These obstacles must satisfy $\underline{\gamma}^J < +\infty$, $-\infty < \overline{\gamma}^J$, and $\underline{\gamma}^J \le \overline{\gamma}^J$.  On the finest-level mesh the discrete boundary data $g_D^J \in \mathcal{V}^J$ is assumed to satisfy $g_D^J = g_D$ on $\partial_D \Omega$---this restricts $g_D$ in \eqref{eq:originalconstraintset} to be piecewise-linear---and $\underline{\gamma}^J \le g_D^J \le \overline{\gamma}^J$ at those vertices of $\mathcal{T}^J$ which lie in $\partial_D \Omega$.  Then we define
\begin{equation}
\mathcal{K}^J = \big\{w\,:\,\underline{\gamma}^J \le w \le \overline{\gamma}^J \text{ on vertices of } \mathcal{T}^J, \, \text{ and } \, w|_{\partial_D\Omega} = g_D^J|_{\partial_D\Omega}\big\} \subset \mathcal{V}^J. \label{eq:fe:fineconstraintset}
\end{equation}
This nonempty, closed, and convex subset is the finest-level \emph{constraint set}.

Let $f^J:\mathcal{K}^J \to (\mathcal{V}^J)'$ be a discretization of $f$ in \eqref{eq:vi}.  Note that, by inclusion \eqref{eq:fe:nestedspaces}, $\ell^J=\ell$ is defined and continuous over $\mathcal{V}^J$.  We seek the FE solution $u^J \in \mathcal{K}^J$ of the finest-level, finite-dimensional, nonlinear VI:
\begin{equation}
\ip{f^J(u^J)}{v-u^J} \ge \ip{\ell^J}{v-u^J} \qquad \text{for all } v\in \cK^J. \label{eq:fe:vi}
\end{equation}
If $f^J$ is continuous and $p$-coercive on $\mathcal{K}^J$ then \eqref{eq:fe:vi} is well-posed.
